To be FAIR: Theory Specification Needs an Update

### 2025-03-02 manuscript.Rmd
update 2025-03-02 14:23:45.942159

## Updating a Theory

@@

The first author was inspired by De Groot as well,
but again specified the empirical cycle differently.
First, notice that the nodes in De Groot's formulation mostly refer to processes.
This invites the question of what the deliverables are in each phase, or in other words: what actually changes when going through the cycle, except the scholar's mind.
In the implementation below, we account for this difference by having the nodes refer to specific deliverables; the edges now refer to processes.
Second, De Groot's distinction between phases of observation, induction, and deduction is not fully congruent with philosophy of science.
<!-- In our representation, -->
<!-- induction is not a separate phase but a mode of reasoning by which specific observations are generalized into theory. -->
<!-- For example, the refutation of a hypothesized effect, -->
<!-- or the serendipitous observation of some pattern in data, -->
<!-- might be a reason to revise or construct theory. -->
<!-- Induction, incidentally, also occurs within the link from prediction to testing: -->
<!-- in the form of the inductive bias of methods used to perform the test, -->
<!-- and auxiliary assumptions that must be made to address remaining theoretical ambiguities. -->
<!-- # And then for example: I understand Aaron's work about inductive bias to be about the link "prediction -> test;", because the test is not identified without making auxiliary assumptions; if the auxiliary assumption is made based on the data, inductive bias is introduced. -->
<!-- ``` -->
Many have argued that observation is theory-laden, and as such, involves induction [@brewerTheoryLadennessObservationTheoryLadenness2001].
Deriving hypotheses from theory is also not purely deductive,
as auxiliary assumptions are often made (which, again, involves induction).
Furthermore, if the testing procedure is not explicitly defined before seeing the data, it incurs some inductive bias as well [@peikertTransparencyOpenScience2023].
With these alterations, we implement the empirical cycle as follows:

>>>>

## Changing a Theory

@@

The first author was inspired by De Groot as well,
but again specified the empirical cycle differently.
First, in De Groot's formulation, each stage describes a process.
This invites the question of what the concrete outcomes of these processes are.
In other words: what actually changes when going through the cycle, except the scholar's mind.
To address this point, the nodes in Van Lissa's specification refer to specific deliverables, whereas the edges connecting the nodes refer to processes acting upon those deliverables, see Figure \@ref(fig:figecs), panel (c).
Second, the processes of induction and deduction are perhaps not as neatly confined to specific phases as De Groot proposed.
<!-- In our representation, -->
<!-- induction is not a separate phase but a mode of reasoning by which specific observations are generalized into theory. -->
<!-- For example, the refutation of a hypothesized effect, -->
<!-- or the serendipitous observation of some pattern in data, -->
<!-- might be a reason to revise or construct theory. -->
<!-- Induction, incidentally, also occurs within the link from prediction to testing: -->
<!-- in the form of the inductive bias of methods used to perform the test, -->
<!-- and auxiliary assumptions that must be made to address remaining theoretical ambiguities. -->
<!-- # And then for example: I understand Aaron's work about inductive bias to be about the link "prediction -> test;", because the test is not identified without making auxiliary assumptions; if the auxiliary assumption is made based on the data, inductive bias is introduced. -->
<!-- ``` -->
For example, if observation is theory-laden, it involves induction [@brewerTheoryLadennessObservationTheoryLadenness2001].
Deriving hypotheses from theory is not purely deductive either,
as auxiliary assumptions are often made, which involves induction (e.g., assuming equal variances across groups because groups often have equal variances).
Furthermore, if the testing procedure is not explicitly defined before seeing the data, it incurs some inductive bias as well [@peikertTransparencyOpenScience2023].
These alterations result in the following implementation of the empirical cycle:
